Finiteness, Cardinality, and Combinatorics in Homotopy Type Theory

### 2020-01-05 paper.tex
Update paper.tex
more

\begin{theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{theorem}
\begin{proof}
  This proof is the composition of two intermediate equivalences: lemmas
  \ref{container-split-enum-equiv} and \ref{container-split-enum-surj}.
  This first of these proves that our definition of finiteness is equivalent to
  another definition which uses lists defined as a \emph{container}, rather than
  the usual inductive type definition.
  The second of these shows that the container-based finiteness predicate is
  equivalent to a split surjection from some \(\mathbf{Fin}\).
\end{proof}

To do so, we will first need to take a small detour to look at Containers.
Containers \cite{abbottContainersConstructingStrictly2005} give us a general way
to construct functors from generic building-blocks.
\begin{definition}[Container]
  A container is a dependent pair of shapes, and positions indexed by those
  shapes.
  \begin{equation}
    \mathbf{Container} = \sum_{(\text{shape} : \text{Type})} , \text{shape} \rightarrow \text{Type}
  \end{equation}
  We ``interpret'' a container into a pair of a specific shape, and a function
  which goes from positions to contents.
  \begin{equation}
    \llbracket S , P \rrbracket = \prod_{(X : \text{Type})} , \sum_{(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{definition}

For instance, the shape of lists is given by \(\mathbb{N}\), and the positions are then
\(\mathbf{Fin}\)s indexed by that \(\mathbb{N}\).
If we swap out wherever we used ``list'' in the definition of split
enumerability with this container-based list, we get an equivalent finiteness
predicate.
\begin{lemma} \label{container-split-enum-equiv}
  The container-based definition of split enumerability is equivalent to the
  inductive type version.
  \begin{equation}
    \sum_{(\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket A)} , \prod_{(x : A)} , x \in xs \simeq
    \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \in xs
  \end{equation}
\end{lemma}
\begin{proof}
  \todo[inline]{Proof}
\end{proof}

This version of split enumerability is equivalent to a split surjection from a
finite prefix of the natural numbers. 
\begin{lemma} \label{container-split-enum-surj}
  Container-defined split enumerability is equivalent to a split surjection from
  some \(\mathbf{Fin}\).
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
\end{proof}
\subsection{Decidable Equality}
\begin{lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
  \ExecuteMetaData[agda/Cardinality/Finite/SplitEnumerable.tex]{discrete}
\end{proof}
This also tells us that any split enumerable type must be a set.
\begin{lemma}
  Any split enumerable type is a set.
\end{lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{lemma}
  \(\bot\) is split enumerable.
\end{lemma}
\begin{proof}
  The support list for \(\bot\) is the empty list.
  The cover proof is present via the principle of explosion.
\end{proof}
\begin{lemma}
  \(\top\) is split enumerable.
\end{lemma}
\begin{proof}
  The support list for \(\top\) is the singleton list containing \(\text{tt}\).
  The cover proof is a constant function which returns an index into the first
  item in the list.
\end{proof}
Both of those cases were relatively simple.
Next, we will look into how to combine proofs of split enumerability.
\begin{theorem}
  Split-enumerability is closed under \(\sum\).
\end{theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \prod_{(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

@@

  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{pi-clos}
\end{proof}
\section{Manifest Enumerability} \label{manifest-enumerability}
As we said in section~\ref{split-enumerable}, we will now retrieve a distinction
between enumerability and manifest Bishop finiteness.
We do so by propositionally truncating the membership proof:
\begin{definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) = \sum_{(\mathit{xs} : \mathbf{List}(A))} , \prod_{(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
  \ExecuteMetaData[agda/Cardinality/Finite/ManifestEnumerable.tex]{def}
\end{definition}

There is another definition, which lies between manifest enumerability and
bishop finiteness, where we would truncate just the equality, but reveal the
position in the membership proof.
\begin{equation}
  \mathcal{E}'(A) = \sum_{(\mathit{xs} : \mathbf{List}(A))} , \prod_{(x : A)} , \sum_{(i : \textbf{Fin}(\text{length}(\mathit{xs})))} , \lVert \mathit{xs} \ind i \equiv x \rVert
\end{equation}
However, this simply amounts to a proof of manifest Bishop finiteness over
set-truncated types, rather than a significantly different construction, so it
does not interest us here.
\subsection{Higher Homotopy Levels}
By hiding the position, we have essentially removed the ``decidable'' component
from split enumerability.
Our predicate now becomes general enough to work with non-sets: we will show
here that the circle is manifestly enumerable.
\begin{theorem}
  The circle \(S^1\) is manifestly enumerable.
\end{theorem}
\begin{proof}
  As the cover proof is a truncated proposition, we need only consider the point
  constructors, making this poof the same as the proof of split enumerability on
  \(\top\).

>>>>

\begin{theorem} \label{split-enum-surj}
  Split enumerability is equivalent to a split surjection from a finite prefix
  of the natural numbers.
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{theorem}
\begin{proof}
  This proof is the composition of two intermediate equivalences: lemmas
  \ref{container-split-enum-equiv} and \ref{container-split-enum-surj}.
  This first of these proves that our definition of finiteness is equivalent to
  another predicate which uses lists defined as a \emph{container}
  (definition~\ref{container-def}).
  The second of these shows that the container-based finiteness predicate is
  equivalent to a split surjection from some \(\mathbf{Fin}\).
\end{proof}

\begin{definition}[Container] \label{container-def}
  A container \cite{abbottContainersConstructingStrictly2005} is a dependent
  pair of shapes, and positions indexed by those shapes. 
  \begin{equation}
    \mathbf{Container} = \sum_{(\text{shape} : \text{Type})} , \text{shape} \rightarrow \text{Type}
  \end{equation}
  We ``interpret'' a container into a pair of a specific shape, and a function
  which goes from positions to contents.
  \begin{equation}
    \llbracket S , P \rrbracket = \prod_{(X : \text{Type})} , \sum_{(s : S)} , \left( P(s) \rightarrow X \right)
  \end{equation}
  Membership of a container can be defined like so:
  \begin{equation}
    x \in \mathit{xs} = \text{fiber}(\text{snd}(\mathit{xs}), x)
  \end{equation}
\end{definition}
We often use containers to define types like lists or trees in a more generic
way.
For instance, while lists would usually be defined like so:
\begin{equation}
  \begin{aligned}
    \mathbf{List}(A) &= \\
      & \left| \; \text{nil} : \mathbf{List}(A)   \right; \\
      & \left| \; \text{cons} : A \rightarrow \mathbf{List}(A) \rightarrow \mathbf{List}(A) \right;
  \end{aligned}
\end{equation}
We could equivalently define them as a container:
\begin{equation}
  \mathbf{List}(A) = \left\llbracket \mathbb{N} , \mathbf{Fin} \right\rrbracket
\end{equation}

The two representations are equivalent, and they lead to two versions of split
enumerability which are equivalent.
\begin{lemma} \label{container-split-enum-equiv}
  The container-based definition of split enumerability is equivalent to the
  inductive type version.
  \begin{equation}
    \sum_{(\mathit{xs} : \llbracket \mathbb{N} , \mathbf{Fin} \rrbracket A)} , \prod_{(x : A)} , x \in xs \simeq
    \sum_{(\mathit{xs} : \textbf{List}(A))} , \prod_{(x : A)} , x \in xs
  \end{equation}
\end{lemma}
\begin{proof}
  \todo[inline]{Proof}
\end{proof}

Our interest in the container-based finiteness predicate comes from the fact
that it is very straightforward to prove it equivalent to a split surjection
from some \(\mathbf{Fin}\).
\begin{lemma} \label{container-split-enum-surj}
  Container-defined split enumerability is equivalent to a split surjection from
  some \(\mathbf{Fin}\).
  \begin{equation}
    \mathcal{E}!(A) \simeq \sum_{(n : \mathbb{N})} , \left( \mathbf{Fin}(n) \twoheadrightarrow ! \; A \right)
  \end{equation}
\end{lemma}
\begin{proof}
  The proof is surprisingly short: after sufficient inlining, it emerges that
  our goal is simply a reassociation.
  \ExecuteMetaData[agda/Cardinality/Container/Finite.tex]{split-surj}
\end{proof}
\subsection{Decidable Equality}
\begin{lemma} \label{split-enum-discrete}
  Any split enumerable type has decidable equality (is discrete).
\end{lemma}
\begin{proof}
  We use a corollary that if there is a split-surjection from \(A\) to \(B\),
  and \(A\) is discrete, then \(B\) is also discrete.
\end{proof}
\begin{lemma}
  Any split enumerable type is a set.
\end{lemma}
\begin{proof}
  By Hedberg's theorem \cite{hedbergCoherenceTheoremMartinLof1998}, since split
  enumerable types have decidable equality
  (proposition~\ref{split-enum-discrete}), they are sets.
\end{proof}
\subsection{Closure}
In this section we will prove closure under various operations for split
enumerable sets.
We are working towards a topos proof, which requires us to prove closure under
a variety of operations: for now, we only have enough machinery to demonstrate
the semiring operations, and dependent sums.
in order to show closure under exponentials (function arrows), we will need an
equivalence with \(\mathbf{Fin}\), which will be provided in
section~\ref{manifest-bishop}.
\begin{lemma}
  \(\bot\) is split enumerable.
\end{lemma}
\begin{proof}
  The support list for \(\bot\) is the empty list.
  The cover proof is present via the principle of explosion.
\end{proof}
\begin{lemma}
  \(\top\) is split enumerable.
\end{lemma}
\begin{proof}
  The support list for \(\top\) is the singleton list containing \(\text{tt}\).
  The cover proof is a constant function which returns an index into the first
  item in the list.
\end{proof}
Both of those cases were relatively simple.
Next, we will look into how to combine proofs of split enumerability.
\begin{theorem}
  Split-enumerability is closed under \(\sum\).
\end{theorem}
\begin{proof}
  Let \(E_A\) be a proof of split enumerability for some type \(A\), and \(E_U\)
  be a function of the type:
  \begin{equation}
    E_U : \prod_{(x : A)} , \mathcal{E}!(U(x))
  \end{equation}
  In other words, a function which returns a proof of split enumerability for
  each member of the family \(U\).

@@

  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{pi-clos}
\end{proof}
\section{Manifest Enumerability} \label{manifest-enumerability}
As we said in section~\ref{split-enumerable}, we will now retrieve a distinction
between enumerability and manifest Bishop finiteness.
We do so by propositionally truncating the membership proof:
\begin{definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) = \sum_{(\mathit{xs} : \mathbf{List}(A))} , \prod_{(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{definition}
\subsection{Higher Homotopy Levels}
By hiding the position, we have essentially removed the ``decidable'' component
from split enumerability.
Our predicate now becomes general enough to work with non-sets: we will show
here that the circle is manifestly enumerable.
\begin{theorem}
  The circle \(S^1\) is manifestly enumerable.
\end{theorem}
\begin{proof}
  As the cover proof is a truncated proposition, we need only consider the point
  constructors, making this poof the same as the proof of split enumerability on
  \(\top\).
